---
title: "LT2 Lab"
output: html_notebook
---

## Packages

```{r}
library(tidyverse)
library(car)
library(lm.beta)
library(mosaic)
library(Hmisc)
library(supernova)
```

## Cleaning: Calculating variables

```{r}
lab <- 
  lab %>% 
  mutate(stress = (STRESS1+STRESS2+STRESS3+STRESS4+STRESS5+STRESS6+STRESS7)/7, sop = (SOP1+SOP2+SOP3+SOP4)/4, spp = (SPP1+SPP2+SPP3+SPP4)/4)
lab # this code calculates the variables
```

## Cleaning: Selecting variables

```{r}
lab <- 
  lab %>% 
  select(sop, spp, stress, HOURS_WORK, HOURS_LV)
lab # I'm just selecting out the key variables here

lab <- # into existing dataframe
lab %>% # from existing daraframe
filter(sop >= "0") # and then filtering just to make sure we have no missing values
```

## Correlations

```{r}
rcorr(as.matrix(lab[,c("sop","spp","stress", "HOURS_WORK", "HOURS_LV")], type="pearson"))
```

## Linear model with sop and spp as predictors

```{r}
lab.model <- ## insert code to build linear model of stress with sop and spp as predictors and save it as object lab.model 
```

## Standardised estimates

```{r}
## insert code for standardised estimates
```

## Bootstrap confidence intervals


```{r}
lab.boot <- Boot(lab.model, f=coef, R = 5000) 
confint(lab.boot, level = .95, type = "norm")
```

## Partitioning variance

```{r}
## insert code to partition lab.model variance using supernova
```


## Adding hours working and hours watching love island to the linear model

```{r}
lab.model2 <- ## build linear model of stress adding HOURS_WORK and HOURS_LV to sop and spp and save it as a new object called lab.model2
```

## Standardised estimates

```{r}
## write code to call standardised estimates using lm.beta
```

## Bootstrap confidence intervals

```{r}
## write the code for bootstrap confidence intervals for lab.model2
```

## Partition variance

```{r}
## insert code to partition lab.model2 variance using supernova
```

