---
title: "LT8 Lab"
output:
  html_document:
    df_print: paged
---

## Packages

```{r eval = FALSE}
install.packages("tidyverse")
install.packages("car")
install.packages("psych")
```


```{r loadlib, echo=T, results='hide', message=F, warning=F}
library(tidyverse)
library(car)
library(psych)
```

## Loading the personality data

Lets load the bigfive dataframe from the survey we just completed.You can do this by clicking the file in the LT8 Lab folder or by trying the code below:

```{r}
bigfive <- read_csv("C:/Users/tc560/Dropbox/Work/LSE/PB130/LT8/Lab/bigfive.csv")
bigfive
```

## Cleaning: Recoding items

The first step when working with psychological test data like this is to identify and tehn recode any reverse items. Recall from the lecture I said some tools have reverse items to mitigate the impact of aquiescence? Well this is true of the Big 5 questionnaire. There are several reverse items for each subscale so we need to recode them so that 5 = 1, 2 = 4, and so on.. Lets do that now with the recode function found in the car package:

```{r}
# first filter out NAs
bigfive <- 
bigfive %>% 
filter(e1 >= "1")

# note I specify car:: here becuase car and tidyverse have a recode function and I want R to use the car recode function
bigfive$a1 <- car::recode(bigfive$a1r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$c2 <- car::recode(bigfive$c2r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$e2 <- car::recode(bigfive$e2r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$n2 <- car::recode(bigfive$n2r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$a3 <- car::recode(bigfive$a3r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$c4 <- car::recode(bigfive$c4r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$e5 <- car::recode(bigfive$e5r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$c5 <- car::recode(bigfive$c5r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$n5 <- car::recode(bigfive$n5r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$a6 <- car::recode(bigfive$a6r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$e7 <- car::recode(bigfive$e7r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$n7 <- car::recode(bigfive$n7r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$o7 <- car::recode(bigfive$o7r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$a8 <- car::recode(bigfive$a8r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$o9 <- car::recode(bigfive$o9r, "1=5; 2=4; 3=3; 4=2; 5=1")
bigfive$c9 <- car::recode(bigfive$c9r, "1=5; 2=4; 3=3; 4=2; 5=1")

# select out the items

bigfive <- select(bigfive, e1, e2, e3, e4, e5, e6, e7, e8, c1, c2, c3, c4, c5, c6, c7, c8, c9, o1, o2, o3, o4, o5, o6, o7, o8, o9, o10, a1, a2, a3, a4, a5, a6, a7, a8, a9, n1, n2, n3, n4, n5, n6, n7, n8)
```

## Cleaning: Reliability analysis

The next step after we have recoded the items is to check the scales are reliable. As we saw in the lecture, we can calculate the reilaibility of the Big 5 scales using the Cronbach's alpha, which is the ratio of variance in the true score to error variance. We saw in the lecture tha tit is calculated as:

r = Vt/V

Where Vt is true score variance, and V is overall variance in the measure.

We can call Cronbach's alpha using the alpha function in the psych package. Lets do that now for each of our subscales; Openness, Extraversion, Neuroticism, Agreeableness, and Concienciousness.

```{r}
extraversion <- select(bigfive, e1, e2, e3, e4, e5, e6, e7, e8)
neuroticism <- select(bigfive, n1, n2, n3, n4, n5, n6, n7, n8)
openness <- select(bigfive, o1, o2, o3, o4, o5, o6, o7, o8, o9, o10)
concienciousness <- select(bigfive, c1, c2, c3, c4, c5, c6, c7, c8, c9)
agreeableness <- select(bigfive, a1, a2, a3, a4, a5, a6, a7, a8, a9)

alpha(neuroticism)
alpha(extraversion)
alpha(openness)
alpha(concienciousness)
alpha(agreeableness)
```

Are the scales reliable? Use Nunally's criteria of 0.70 to answer.

## Cleaning: Calculating variables

Once we have established that the Big 5 scales are reliable, we can move to aggregating the items to creating the variables for each sub-scale. We can easily do this with the mutate function, as we saw at the beginning of the first semester.

```{r}
bigfive <- bigfive %>%
  mutate(neuroticism = (n1 + n2 + n3 + n4 + n5 + n6 + n7 + n8)/8)
bigfive <- bigfive %>%
  mutate(openness = (o1 + o2 + o3 + o4 + o5 + o6 + o7 + o8 + o9 + o10)/10)
bigfive <- bigfive %>%
  mutate(extraversion = (e1 + e2 + e3 + e4 + e5 + e6 + e7 + e8)/8)
bigfive <- bigfive %>%
  mutate(concienciousness = (c1 + c2 + c3 + c4 + c5 + c6 + c7 + c8 + c9)/9)
bigfive <- bigfive %>%
  mutate(agreeableness = (a1 + a2 + a3 + a4 + a5 + a6 + a7 + a8 + a9)/9)
```

## Activity: Plot a bar graph showing the mean scores for the personality dimensions


```{r}
# First create dataframe with means
mean_score <- c(mean(bigfive$neuroticism), mean(bigfive$openness), mean(bigfive$extraversion), mean(bigfive$concienciousness), mean(bigfive$agreeableness))
personality <- c("neuroticism", "openness", "extraversion", "concienciousness", "agreeableness")
data <- data_frame(mean_score, personality)

## Now create the bar chart

```

